---
title: "GRUK_effekt_circles_2024"
format: html
editor: visual
---

## Introduction

This script creates figures for the observations in circles. In addition, it contains some simple statistical tests.

## Language settings

```{r}
Sys.setlocale("LC_CTYPE", "nb_NO.UTF-8")
```

## Install packages

```{r}

library(tidyverse)
library(readxl)
library(forcats)
library(ggpubr)
```

## Import data

The circle data available in datafile prepared by M. Evju.

```{r}

# Define the file path to your Excel file
file_path <- here("Apen-grunnland-kalkmark/data/GRUK_nytt_format_copy.xlsx")

# Read each sheet and add a month column
dat_circles <- read_excel(file_path, sheet = "circles")

```

## Bleikøya

Let's create a figure that shows the change in cover of:

-   invasive species

-   woody plants in field layer

-   shrub layer

-   canvas/tarp

```{r}

p_bleikøya_FA <- dat_circles %>% 
  filter(Lokalitet == "Bleikøya") %>% 
  ggplot(aes(x = Tid, y = Circle.invasive.species))+
  geom_boxplot() +
  theme_classic() +
  labs(x = "Tid", y = "Fremmedarter (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100))

p_bleikøya_woody_ground <- p_bleikøya_FA + aes(x = Tid, y = Circle.woody.field.layer) +
  labs(y = "Vedplanter i feltsjikt (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,50))

p_bleikøya_shrubs <- p_bleikøya_FA + aes(x = Tid, y = Circle.shrub.layer) +
  labs(y = "Busksjikt (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,50))

p_bleikøya_canvas <- p_bleikøya_FA + aes(x = Tid, y = as.numeric(Circle.canvas)) +
  labs(y = "Duk (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,50))

#Create one combined plot

p_bleikøya_all <- ggarrange(p_bleikøya_canvas, p_bleikøya_FA, p_bleikøya_woody_ground, p_bleikøya_shrubs, labels = c("a", "b", "c", "d"), common.legend = TRUE, legend = "none", ncol = 2, nrow = 2)

```

## Padda

For Padda, we only consider the plots that are in GRUK and ignore the ones that are located in the forest.

```{r}

plot_padda <- c("P1", "P10", "P15", "P2", "P3", "P4", "P5", "P5", "P6", "P9")

p_padda_FA <- dat_circles %>% 
  filter(Rute %in% plot_padda) %>% 
  ggplot(aes(x = Tid, y = Circle.invasive.species))+
  geom_boxplot() +
  theme_classic() +
  labs(x = "Tid", y = "Fremmedarter (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100))

p_padda_PA <- p_padda_FA + aes(x = Tid, y = Circle.problematic.species)+
  labs(y = "Problemarter (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100))#her er sneglebelg med

p_padda_woody_ground <- p_padda_FA + aes(x = Tid, y = Circle.woody.field.layer) +
  labs(y = "Vedplanter i feltsjikt (%)")+
  scale_y_continuous(expand = c(0, 0), limits = c(0,100))

p_padda_shrubs <- p_padda_FA + aes(x = Tid, y = Circle.shrub.layer) +
  labs(y = "Busksjikt (%)")
  scale_y_continuous(expand = c(0, 0), limits = c(0,100))
  
p_padda_slitasje <- p_padda_FA + aes(x = Tid, y = as.numeric(Circle.erosion)) +
  labs(y = "Slitasje (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,50))

p_padda_vehicletracks <- p_padda_FA + aes(x = Tid, y = as.numeric(Circle.vehicle.tracks)) +
  labs(y = "Kjørespor (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100))

p_padda_all <- ggarrange(p_padda_FA, p_padda_PA, p_padda_woody_ground, p_padda_shrubs, p_padda_vehicletracks, p_padda_slitasje, labels = c("a", "b", "c", "d", "e", "f"), common.legend = TRUE, legend = "none", ncol = 3, nrow = 2)


```

## Torvøya

```{r}

p_torvøya_FA <- dat_circles %>% 
  filter(Lokalitet == "Torvøya") %>% 
  ggplot(aes(x = Tid, y = Circle.invasive.species))+
  geom_boxplot() +
  theme_classic() +
  labs(x = "Tid", y = "Fremmedarter (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100))

p_torvøya_woody_ground <- p_torvøya_FA + aes(x = Tid, y = Circle.woody.field.layer) +
  labs(y = "Vedplanter i feltsjikt (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,50))

p_torvøya_shrubs <- p_torvøya_FA + aes(x = Tid, y = Circle.shrub.layer) +
  labs(y = "Busksjikt (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,50))

p_torvøya_all <- ggarrange(p_torvøya_FA, p_torvøya_woody_ground, p_torvøya_shrubs, labels = c("a", "b", "c"), common.legend = TRUE, legend = "none", ncol = 3, nrow = 1)


```

## **Malmøya**

```{r}
p_malmøya_FA <- dat_circles %>% 
  filter(Lokalitet == "Malmøya") %>% 
  ggplot(aes(x = Tid, y = Circle.invasive.species))+
  geom_boxplot() +
  theme_classic() +
  labs(x = "Tid", y = "Fremmedarter (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,50))

p_malmøya_PA <- p_malmøya_FA + aes(x = Tid, y = Circle.problematic.species)+
  labs(y = "Problemarter (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100))#her er syrin med

p_malmøya_woody_ground <- p_malmøya_FA + aes(x = Tid, y = Circle.woody.field.layer) +
  labs(y = "Vedplanter i feltsjikt (%)")+
  scale_y_continuous(expand = c(0, 0), limits = c(0,100))

p_malmøya_shrubs <- p_malmøya_FA + aes(x = Tid, y = Circle.shrub.layer) +
  labs(y = "Busksjikt (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100))
  
p_malmøya_trees <- p_malmøya_FA + aes(x = Tid, y = Circle.tree.layer) +
  labs(y = "Tresjikt (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100))
  
p_malmøya_slitasje <- p_malmøya_FA + aes(x = Tid, y = as.numeric(Circle.erosion)) +
  labs(y = "Slitasje (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100))


p_malmøya_all <- ggarrange(p_malmøya_FA, p_malmøya_PA, p_malmøya_woody_ground, p_malmøya_shrubs, p_malmøya_trees, p_malmøya_slitasje, labels = c("a", "b", "c", "d", "e", "f"), common.legend = TRUE, legend = "none", ncol = 3, nrow = 2)


```

## Husbergøya

```{r}
p_husbergøya_FA <- dat_circles %>% 
  filter(Lokalitet == "Husbergøya") %>% 
  ggplot(aes(x = Tid, y = Circle.invasive.species))+
  geom_boxplot() +
  theme_classic() +
  labs(x = "Tid", y = "Fremmedarter (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,50))

p_husbergøya_PA <- p_husbergøya_FA + aes(x = Tid, y = Circle.problematic.species)+
  labs(y = "Problemarter (%)")

p_husbergøya_woody_ground <- p_husbergøya_FA + aes(x = Tid, y = Circle.woody.field.layer) +
  labs(y = "Vedplanter i feltsjikt (%)")

p_husbergøya_shrubs <- p_husbergøya_FA + aes(x = Tid, y = Circle.shrub.layer) +
  labs(y = "Busksjikt (%)")
  
p_husbergøya_trees <- p_husbergøya_FA + aes(x = Tid, y = Circle.tree.layer) +
  labs(y = "Tresjikt (%)")
  
p_husbergøya_slitasje <- p_husbergøya_FA + aes(x = Tid, y = as.numeric(Circle.erosion)) +
  labs(y = "Slitasje (%)") 


p_husbergøya_all <- ggarrange(p_husbergøya_FA, p_husbergøya_PA, p_husbergøya_woody_ground, p_husbergøya_shrubs, p_husbergøya_trees, p_husbergøya_slitasje, labels = c("a", "b", "c", "d", "e", "f"), common.legend = TRUE, legend = "none", ncol = 3, nrow = 2)
```
